Pulling oral and IV data from CvTdb

### 2025-04-17 data-raw/pulling_oral_iv_data_cvtdb.Rmd
Secondary update to cvt object and related files.

      diff_columns <- function(x, y) {
        out_list <- character()
        for (ix in names(x)) {
          for (iy in names(y)) {
            if (identical(x[[c(key_id,ix)]], y[[c(key_id, iy)]])) {
              out_list <- c(out_list, paste0(ix, " -> ", iy))
            }
          }
        }
        return(out_list)
      }

      renamed_cols <- diff_columns(common_row_cvt[diff_col_sub],
                                   common_row_new[diff_col_add])
      if (length(renamed_cols) > 0) {
        msg_to_write <- c(msg_to_write,
                          "Renamed columns (non-exhaustive):",
                          renamed_cols)
      } else {
        msg_to_write <- c(msg_to_write,
                          "No renamed columns found.")
      }
    }
    if (length(diff_col_add) > 0) {
      msg_to_write <- c(msg_to_write,
                        "--------------",
                        "Columns added:",
                        diff_col_add)
    }
    if (length(diff_col_sub) > 0) {
      msg_to_write <- c(msg_to_write,
                        "______________",
                        "Columns removed (from previous data):",
                        diff_col_sub)
    }

@@

# Harmonize values for easier analysis
cvt_df <- sql_query %>%
  mutate(
    across(ends_with("id"), as.character),
    # Note that sometimes fk_test_chemical_id is NA
    across(all_of(c("time_original",
                    "time_hr",
                    "conc_original",
                    "conc_sd_original",
                    "conc",
                    "conc_sd",
                    "dose_level_original",
                    "dose_level_normalized",
                    "loq")), as.numeric),
    # conc, conc_sd_original, and conc_original
    # have NA/NE/ND values that are coerced to NAs
    year = as.character(year),
    species = tolower(species),
    sex = tolower(sex),
    analyte_name_original = tolower(analyte_name_original),
    dose_volume_units = stringr::str_extract(dose_volume,
                                    pattern = "(?<=[:space:]).+"),
    dose_volume = as.numeric(stringr::str_extract(dose_volume,
                                         pattern = "^(\\d|\\.)*")),
    dose_duration_units = stringr::str_extract(dose_duration,
                                      pattern = "(?<=[space]).+"),
    dose_duration = as.numeric(stringr::str_extract(dose_duration,
                                pattern = "^\\d")),
    radiolabeled = ifelse(radiolabeled == 1, TRUE, FALSE),
    n_subjects_normalized = ifelse(n_subjects_in_series %in% "NR", NA_real_,
      as.numeric(stringr::str_extract(n_subjects_in_series,
                                                 pattern = "\\d+")))
  )

@@

# There are some missing DTXSIDs that we need to fix
# cvt_df %>%
#   distinct(analyte_dtxsid, analyte_name_original, fk_analyzed_chemical_id) %>%
#   filter(is.na(analyte_dtxsid)) %>% nrow()

@@

dq_cols <- dtxsid_query %>%
  select("analyte_dtxsid2" = "dsstox_substance_id",
         "fk_analyzed_chemical_id" = "id",
         "analyte_name_original" = "chemical_name_original") %>%
  mutate(analyte_name_original =  tolower(analyte_name_original)) %>%
  distinct()

dq_cols <- dq_cols %>% distinct(analyte_dtxsid2, analyte_name_original) %>%
  group_by(analyte_name_original) %>%
  summarize(analyte_dtxsid2 = if_else(any(complete.cases(analyte_dtxsid2)),
                                      first(analyte_dtxsid2[!is.na(analyte_dtxsid2)]),
                                      NA)) %>%
  left_join(dq_cols %>% select(-analyte_dtxsid2),
            join_by(analyte_name_original)) %>%
  mutate(fk_analyzed_chemical_id = as.character(fk_analyzed_chemical_id))

@@

study_query <- query_db(query = "SELECT distinct * FROM studies",
         schema = "cvt")

# Here we join to make sure all chemical ids are represented
full_set <- cvt_df %>%
  left_join(dq_cols, 
            by = join_by("fk_analyzed_chemical_id", "analyte_name_original")) %>%
  mutate(analyte_dtxsid = coalesce(analyte_dtxsid, analyte_dtxsid2)) %>%
  select(-analyte_dtxsid2)

cvt_df <- full_set
quick_stats(cvt_df)

# For figuring out whether there was a change from the original units to reported conc
# And only keep non-radiolabeling experiments, where analyte_dtxsid is known
# and where analyzed chemical is equivalent to the dosed chemical
cvt_df <- cvt_df %>% mutate(conc_unit_norm_factor = signif(conc/conc_original, 1),
                            .after = conc_units_normalized) %>%
  filter(!(radiolabeled %in% TRUE),
         !is.na(analyte_dtxsid),
         fk_analyzed_chemical_id == fk_dosed_chemical_id)

quick_message(text = c("Filtering out:",
                       "Missing {.code analyte_dtxsid}",
                       "Radiolabeling experiments",
                       "Data where the analyzed chemicals and dosed chemicals are different"))
diff_log_cvt(cvt_df)
```

@@

# Filtering out entries with no concentration information available
cvt_df <- cvt_df %>%
  filter(!(is.na(conc) & is.na(conc_original))) %>%
  group_by(fk_study_id) %>%
  filter(n() > 1) %>%
  ungroup()
###
```

@@

+  Convert original dose units for CEBS study _S0916_ to mg/kg.  
+  ~~Convert original dose units for CEBS study _S0624_ to mg/kg.~~ Data got corrected in CEBS, trusting that.  
+  Input the LOQ for studies _S0976_.  
+  Infer the LOQ for study _S0328_ as 1, given the values near it deemed above and below LOQ.

```{r specific_data_validation}
# Looking for suspect dose levels
quick_message(paste0("Looking at dose units not equivalent to mg/kg where ",
                     "dose_level_original and dose_level_normalized ",
                     "{.emph remain the same}"))
cvt_df %>% filter(!(dose_level_units_original %in% equiv_mgkg),
                  dose_level_normalized == dose_level_original) %>%
  distinct(fk_study_id, 
           dose_level_original, dose_level_normalized,
           dose_level_units_original,
           pmid,
           other_study_identifier)

@@

quick_message("CEBS datasets S0976 and S0328 contain data on LOQ or ELOQ")
# For study S0976, Caffeine, L-ephedrine, and psuedoephedrine 
quick_message(c(
  "Data Modification:",
  "Adjusting LOQ values for L-ephedrine to 3.39 as stated in CEBS study S0976.",
  "Set LOQ units to ng/mL as detailed in the same CEBS study.",
  "Caffeine and Pseudoephedine have been removed as of April 2025."
))
# Here I create new columns for loq and loq_units b/c I have modified them.
cvt_df <- cvt_df %>%
  mutate(invivPK_loq = if_else(
    other_study_identifier %in% "S0976",
    case_when(
      # analyte_name_original %in% "caffeine" ~ 4.00,
      analyte_name_original %in% "l-ephedrine" ~ 3.39,
      # analyte_name_original %in% "pseudoephedrine" ~ 4.43
    ),
    loq),
    invivPK_loq_units = if_else(
      other_study_identifier %in% "S0976",
      "ng/mL",
      loq_units
    ))

@@

# quick_stats(cvt_df)
diff_log_cvt(cvt_df)
```

@@

# quick_stats(cvt_df)
quick_message("Uniqe LOQ units:")
table(cvt_df$invivPK_loq_units)
diff_log_cvt(cvt_df)

@@

quick_message("Are there chemicals with very few (less than 5) observations still?")
cvt_df %>% count(analyte_dtxsid) %>% arrange(n) %>% head()

quick_message(c(
  "Filtering Data:",
  "Removing DTXSID6038299 due to insufficient data",
  "Removing all data points that still have recorded concentrations of zero"
))
cvt_df <- cvt_df %>%
  dplyr::filter(!(analyte_dtxsid %in% "DTXSID6038299"))

@@

glimpse(cvt_df, width = 80)
quick_stats(cvt_df)
diff_log_cvt(cvt_df, 
             notes = "Relaxed conc > 0 filter to conc >= 0.")

@@

```{r saving_data, include=FALSE, eval=FALSE}
cvt_date <- Sys.Date()
cat(diff_log_cvt(cvt_df,
                 notes = "Relaxed conc > 0 filter to conc >= 0."),
    file = "data-raw/cvt_changelog.txt",
    append = TRUE)

>>>>

      diff_columns <- function(x, y) {
        out_list <- character()
        for (ix in setdiff(names(x), key_id)) {
          for (iy in setdiff(names(y), key_id)) {
            tmp <- merge(x[c(key_id,ix)], y[c(key_id, iy)], by=key_id)
            if (identical(tmp[[2]], tmp[[3]])) {
              out_list <- c(out_list, paste0(ix, " -> ", iy))
            }
          }
        }
        return(out_list)
      }

      renamed_cols <- diff_columns(common_row_cvt[c(key_id,diff_col_sub)],
                                   common_row_new[c(key_id, diff_col_add)])
      if (length(renamed_cols) > 0) {
        msg_to_write <- c(msg_to_write,
                          "Renamed columns (non-exhaustive):",
                          renamed_cols)
      } else {
        msg_to_write <- c(msg_to_write,
                          "No renamed columns found.")
      }
    }
    if (length(diff_col_add) > 0) {
      msg_to_write <- c(msg_to_write,
                        "--------------",
                        "Columns added:",
                        diff_col_add)
    }
    if (length(diff_col_sub) > 0) {
      msg_to_write <- c(msg_to_write,
                        "______________",
                        "Columns removed (from previous data):",
                        diff_col_sub)
    }

@@

# Unfortunately, there are some "fancy" or curly quotes that need to be replaced
# with "straight" quotes to be more machine read-able
force_straight_quotes <- function(x) {
  temp <- gsub("(’|′)", "'", x)
  temp <- gsub("(, | ,)", ",", temp)
  return(temp)
}
force_straight_quotes("3,3’,4,4’-tetrachloro-5-biphenylol")

# Harmonize values for easier analysis
cvt_df <- sql_query %>%
  mutate(
    across(ends_with("id"), as.character),
    # Note that sometimes fk_test_chemical_id is NA
    across(all_of(c("time_original",
                    "time_hr",
                    "conc_original",
                    "conc_sd_original",
                    "conc",
                    "conc_sd",
                    "dose_level_original",
                    "dose_level_normalized",
                    "loq")), as.numeric),
    # conc, conc_sd_original, and conc_original
    # have NA/NE/ND values that are coerced to NAs
    year = as.character(year),
    species = tolower(species),
    sex = tolower(sex),
    analyte_name_original = force_straight_quotes(tolower(analyte_name_original)),
    analyte_name_secondary_original = force_straight_quotes(tolower(
      analyte_name_secondary_original
      )),
    dose_volume_units = stringr::str_extract(dose_volume,
                                    pattern = "(?<=[:space:]).+"),
    dose_volume = as.numeric(stringr::str_extract(dose_volume,
                                         pattern = "^(\\d|\\.)*")),
    dose_duration_units = stringr::str_extract(dose_duration,
                                      pattern = "(?<=[space]).+"),
    dose_duration = as.numeric(stringr::str_extract(dose_duration,
                                pattern = "^\\d")),
    radiolabeled = ifelse(is.na(radiolabeled), FALSE, as.logical(radiolabeled)),
    n_subjects_normalized = ifelse(n_subjects_in_series %in% "NR", NA_real_,
      as.numeric(stringr::str_extract(n_subjects_in_series,
                                                 pattern = "\\d+")))
  )

@@

# There are some missing DTXSIDs that we need to fix
cvt_df %>% 
    filter(is.na(analyte_dtxsid),
           !(radiolabeled %in% TRUE)) %>%
  distinct(fk_analyzed_chemical_id,fk_dosed_chemical_id,
           analyte_dtxsid, analyte_name_original,
           analyte_name_secondary_original) %>%
  View("NA_dtxsid")

@@

dtxsid_query <- dtxsid_query %>%
  filter(id %in% as.numeric(
    union(cvt_df$fk_analyzed_chemical_id, cvt_df$fk_dosed_chemical_id))
  ) %>%
  mutate(id = as.character(id),
         chemical_name_original = force_straight_quotes(tolower(
           chemical_name_original
         )),
         chemical_name_secondary_original = force_straight_quotes(tolower(
           chemical_name_secondary_original
         ))
  )

cvt_df %>% 
  filter(fk_analyzed_chemical_id != fk_dosed_chemical_id,
         !(radiolabeled %in% TRUE)) %>%
  distinct(fk_analyzed_chemical_id,fk_dosed_chemical_id,fk_test_chemical_id) %>%
  left_join(dtxsid_query %>%
              select("fk_analyzed_chemical_id" = "id",
                     "analyzed_chemical_name" = "chemical_name_original")) %>%
  left_join(dtxsid_query %>%
              select("fk_dosed_chemical_id" = "id",
                     "dosed_chemical_name" = "chemical_name_original")) %>%
  View("errors")

# Rather than hope most of these chemicals have dtxsids within CvTdb,
# I cross-referenced it with the CompTox database batch search.
dtxsid_batch_idless <- readr::read_csv(
  "data-raw/CCD-Batch-Search_2025-04-17_DTXSIDless.csv",
  na = "N/A"
) %>% filter(!is.na(DTXSID)) %>%
  select("dsstox_substance_id" = "DTXSID",
         "chemical_name_original" = "INPUT") %>%
  distinct()

quick_message(c(
  "Erroneous DTXSID assignments (identified via duplication):",
  "perfluorohexanesulfonate == DTXSID80873012",
  "tacrolimus == DTXSID5046354",
  "hippuric acid == DTXSID9046073"
))

dtxsid_batch_idless <- dtxsid_batch_idless %>%
  mutate(dsstox_substance_id = case_when(
    chemical_name_original == "perfluorohexanesulfonate" ~ "DTXSID80873012",
    chemical_name_original == "tacrolimus" ~ "DTXSID5046354",
    chemical_name_original == "hippuric acid" ~ "DTXSID9046073",
    .default = dsstox_substance_id
  )) %>% distinct()

# Fixing the missing DTXSID to prevent premature filtering
cvt_df <- cvt_df %>%
  left_join(
    dtxsid_batch_idless,
    by = join_by(analyte_name_original == chemical_name_original)
  ) %>%
  mutate(analyte_dtxsid = coalesce(analyte_dtxsid, dsstox_substance_id)) %>%
  select(!dsstox_substance_id)

quick_stats(cvt_df)

# Now we need to pre-emptively address the chemicals that have different
# dosed_id and analyzed_ids but are actually the same chemical
quick_message(c(
  "The following have resolvable errors:",
  "If dosed chemical id is 47, change this to 46 to match analyte id. tamoxifen, salt dosed.",
  "If dosed chemical id is 893, change this to 929 to match analyte id. trichloroethylene, duplicate.",
  "If dosed chemical id is 671, change to 672 to match analyte id. PFHS, salt dosed.",
  "If dosed chemical id is 673, change to 675 to match analyte id. PFNA, capitalization.",
  "If dosed chemical id is 674, change to 676 to match analyte id, PFDA, capitalization."
))

cvt_df <- cvt_df %>%
  rowwise() %>%
  mutate(fk_dosed_chemical_id = case_when(
    fk_dosed_chemical_id == "47" &&
      fk_analyzed_chemical_id == "46" ~ "46",
    fk_dosed_chemical_id == "893" &&
      fk_analyzed_chemical_id == "929" ~ "929",
    fk_dosed_chemical_id == "671" &&
      fk_analyzed_chemical_id == "672" ~ "672",
    fk_dosed_chemical_id == "673" &&
      fk_analyzed_chemical_id == "673" ~ "673",
    fk_dosed_chemical_id == "674" &&
      fk_analyzed_chemical_id == "676" ~ "676",
    .default = fk_dosed_chemical_id
  )) %>%
  ungroup()

@@

# study_query <- query_db(query = "SELECT distinct * FROM studies",
#          schema = "cvt")

quick_stats(cvt_df)

# For figuring out whether there was a change from the original units to reported conc
# And only keep non-radiolabeling experiments, where analyte_dtxsid is known
# and where analyzed chemical is equivalent to the dosed chemical
cvt_df <- cvt_df %>% mutate(conc_unit_norm_factor = signif(conc/conc_original, 1),
                            .after = conc_units_normalized)

cat("Current # chemicals: ", length(unique(cvt_df$analyte_dtxsid)), "\n")
cat("Number of chemicals where analyte_dtxsid is NA: ",
    sum(is.na(unique(cvt_df[c("fk_analyzed_chemical_id",
                               "analyte_dtxsid")])$analyte_dtxsid)),
    "\n")

cat("Number of radiolabelling experiments: ",
    sum(distinct(cvt_df, analyte_dtxsid, radiolabeled) %>% pull(radiolabeled) %>%
          {. %in% TRUE}
        )
    )
distinct(cvt_df, analyte_dtxsid, analyte_name_original, radiolabeled) %>%
  filter(radiolabeled %in% TRUE, !is.na(analyte_dtxsid))

cat("Number of chemicals with non-matching ids ",
    cvt_df %>%
      filter(!is.na(analyte_dtxsid),
             !(radiolabeled %in% TRUE),
             fk_analyzed_chemical_id != fk_dosed_chemical_id) %>%
      distinct(analyte_dtxsid) %>%
      NROW()
)

cvt_df <- cvt_df %>%
  filter(!(radiolabeled %in% TRUE),
         !is.na(analyte_dtxsid),
         (fk_analyzed_chemical_id == fk_dosed_chemical_id)) %>%
  distinct()

quick_message(text = c("Filtering out:",
                       "Missing {.code analyte_dtxsid}",
                       "Radiolabeling experiments",
                       "Data where the analyzed chemicals and dosed chemicals are different"))
quick_stats(cvt_df)
diff_log_cvt(cvt_df)
```

@@

# Filtering out entries with no concentration information available
cvt_df <- cvt_df %>%
  filter(!(is.na(conc) & is.na(conc_original))) %>%
  group_by(fk_study_id) %>%
  filter(n() > 1) %>%
  ungroup()
###
quick_stats(cvt_df)
```

@@

+  Convert original dose units for CEBS study _S0916_ to mg/kg.  
+  Convert original dose units for CEBS study _S0624_ to mg/kg. Data got corrected in CEBS, trusting that.  
+  Input the LOQ for studies _S0976_.  
+  Infer the LOQ for study _S0328_ as 1, given the values near it deemed above and below LOQ.

```{r specific_data_validation}
# Looking for suspect dose levels
quick_message(paste0("Looking at dose units not equivalent to mg/kg where ",
                     "dose_level_original and dose_level_normalized ",
                     "{.emph remain the same}"))

cvt_df %>% filter(!(dose_level_units_original %in% equiv_mgkg),
                  dose_level_normalized == dose_level_original) %>%
  distinct(fk_study_id, 
           dose_level_original, dose_level_normalized,
           dose_level_units_original,
           pmid,
           other_study_identifier)

@@

quick_message("CEBS datasets S0976 and S0328 contain data on LOQ or ELOQ")
# For study S0976, Caffeine, L-ephedrine, and psuedoephedrine 
quick_message(c(
  "Data Modification:",
  "Adjusting LOQ values for L-ephedrine to 3.39 as stated in CEBS study S0976.",
  "Set LOQ units to ng/mL as detailed in the same CEBS study.",
  "S0976 has been removed as of April 2025."
))
# Here I create new columns for loq and loq_units b/c I have modified them.
# cvt_df <- cvt_df %>%
#   mutate(invivPK_loq = if_else(
#     other_study_identifier %in% "S0976",
#     case_when(
#       analyte_name_original %in% "caffeine" ~ 4.00,
#       analyte_name_original %in% "l-ephedrine" ~ 3.39,
#       analyte_name_original %in% "pseudoephedrine" ~ 4.43
#     ),
#     loq),
#     invivPK_loq_units = if_else(
#       other_study_identifier %in% "S0976",
#       "ng/mL",
#       loq_units
#     ))

@@

quick_stats(cvt_df)
diff_log_cvt(cvt_df)
```

@@

# quick_stats(cvt_df)
quick_message("Uniqe LOQ units:")
table(cvt_df$invivPK_loq_units)
diff_log_cvt(cvt_df)
quick_stats(cvt_df)

@@

quick_message("Are there chemicals with very few (less than 5) observations still?")
cvt_df %>% count(analyte_dtxsid) %>% arrange(n) %>% head()
# This has been resolved now
# quick_message(c(
#   "Filtering Data:",
#   "Removing DTXSID6038299 due to insufficient data",
#   "Removing all data points that still have recorded concentrations of zero"
# ))
# cvt_df <- cvt_df %>%
#   dplyr::filter(!(analyte_dtxsid %in% "DTXSID6038299"))

@@

glimpse(cvt_df)
quick_stats(cvt_df)

diff_log_cvt(cvt_df, 
             notes = c("Used CompTox name - DTXSID matching to fill in NAs.",
                       "Fixed names that used non-standard single quotes.")
)

@@

```{r saving_data, include=FALSE, eval=FALSE}
cvt_date <- Sys.Date()
cat(diff_log_cvt(cvt_df, 
             notes = c("Used CompTox name - DTXSID matching to fill in NAs.",
                       "Fixed names that used non-standard single quotes.")
),
    file = "data-raw/cvt_changelog.txt",
    append = TRUE)
